Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2021-04-05 bbbq_article.tex
Review paper on 2021-04-05, progress #212

Cytolytic T cell responses are predicted to be biased towards membrane proteins. 
Because the peptide-binding grooves of most haplotypes 
of histocompatibility complex class I (MHC-I) are relatively hydrophobic, 
peptide fragments derived from human transmembrane helices (TMHs) 
are predicted to be presented more often than expected
based on their abundance in the proteome.
We show that this over-presentation of TMH-derived peptides is  general, 
as it is predicted for diverse bacteria and viruses 
and for both MHC-I and MHC-II.  However, the physiological reason of why membrane proteins might be 
over-presented is unclear
In this study, we show that TMHs are evolutionarily more conserved, 
as single nucleotide polymorphisms (SNPs) 
are present relatively less frequent in TMH-coding chromosomal regions 
compared to regions coding for extracellular and cytosolic protein regions. 
Thus, our findings suggest that T cells might respond 
more to membrane proteins, because these are evolutionary more conserved.
We speculate that TMHs therefor might be less prone to escape mutations 
that enable pathogens to evade T cell responses.

@@

The MHC proteins are heterodimeric proteins encoded by the
HLA (Human Leukocyte Antigens) genes.
In humans, the peptide-binding groove of MHC-I is made by the alpha chain, and there are three forms of alpha chains called HLA-A, HLA-B and HLA-C. 
For MHC-II, both the alpha and the beta chains contribute to the binding groove, and there are three major forms as well, called HLA-DR, HLA-DQ and HLA-DP.
Each MHC complex can only bind a subset of all possible peptides.
For example, HLA-A and HLA-B have no overlap in which
peptides they bind (\cite{lund2004definition})
The HLA genes of humans are highly polymorphic, with hundreds 
to thousands of different alleles, 
and each different HLA allele is called 
an MHC haplotype (\cite{marsh2010nomenclature}).

@@

While humans only express a limited number of MHC haplotypes, and therefore any individual's immune system detects only a fraction of all possible
peptide fragments, the number of pathogenic peptides that can be detected is increased at the population level, because of the multiple and highly polymorphic MHC genes.
It is believed that this improves immunity at the population level, 
as mutations in a protein that disrupt a particular MHC presentation, 
so-called escape mutations, 
will not affect MHC presentation for all haplotypes (\cite{sommer2005importance}).
% Do not give the number of MHC molecules, as is is more complex, 
% as the alpha chains coded by the maternal chromosome 
% of HLA-DQ, DR, DP can heterodimerize with the beta chains coded 
% by the paternal chromosome and vice versa

@@

Many studies are aimed at determining which peptides are presented in MHC 
and will result in an immune response, 
as this will for instance aid the design of vaccines. 
These studies have led to the development 
of  prediction algorithms 
that allow for very reliable in silico predictions 
of the binding affinities of peptides (\cite{larsen2010identification,schellens2008unanticipated,tang2011genome}).
For example, \cite{tang2011genome} found that, 
of the 432 peptides that were predicted to bind to MHC,
86\% were confirmed to do so in vivo.

@@

Transmembrane helices (TMHs) are hydrophobic 
as they need to span the hydrophobic lipid bilayer of cellular membranes.
They consist of an alpha helix of on average 23 amino acids in length. TMHs can also be predicted from a protein sequence 
with high accuracy by bioinformatics approaches (\cite{krogh2001predicting,bianchi2017,kall2004combined,arai2004conpred,jones2007improving,klammer2009metatm,wang2019efficient}),
for example, \cite{jones2007improving} found that,
from 184 TMPs with known topology, 
80\% of the TMH predictions replicated this finding.

TMHs are common structures in the proteins of humans and microbes. 
Different TMH prediction tools estimate
that 15-39\% of the human proteome 
contain at least one TMH (\cite{ahram2006estimation}).
However, the physiological reason why peptides derived from TMHs 
would be presented more often than peptides 
stemming from soluble protein regions is unknown. 
We hypothesized is that the presentation of 
TMH residues is evolutionary selected for, 
because TMHs are less prone to undergo escape mutation. 
One reason to expect such a reduced 
variability (and hence evolutionary conservation) in TMHs, 
is that these are restricted in their evolution 
by the functional requirement to span a lipid bilayer. 
Due to this requirement, 
many of the amino acids genuinely present in TMH 
are limited to the ones with hydrophobic side chains (\cite{jones1994model}).
Therefore, we speculated that the TMHs of pathogens 
might have a lower chance to develop escape mutations, 
as many mutations will result in a dysfunctional TMH 
and render the protein inactive.

This study has two objectives. First, we aimed to generalize our findings by predicting the presentation of peptides from different kingdoms of life and for both MHC-I and -II. From these in silico predictions, we conclude that TMH-derived
epitopes are presented more often than expected by chance,
in a human, viral and bacterial proteome, and for most haplotypes of both MHC-I and II. We confirmed this over-presentation of TMH-derived peptides by re-analysis of a peptide elution study. Second, we tested our hypothesis that TMHs
are more evolutionarily conserved than  solvent-exposed protein regions. Our analysis of human
single nucleotide polymorphisms (SNPs) showed that random point mutations are indeed less likely
to occur within TMHs. These findings strengthen the emerging notion that TMHs are important for the T cell branch of the adaptive immune system, and hence are of  
overlooked importance in vaccine development.

@@

The MHC-I molecule presents epitopes of shorter length than MHC-II.
We expect this to be of no influence in the presentation level
of TMH-derived epitopes, for each of the three proteomes.
We verified this, by normalizing the percentage of TMH-derived epitopes found
to the percentage of TMH-derived epitopes expected, 
where a value of $1.0$ denotes that the percentage of TMH-derived epitopes found
matches the percentage as expected by chance.

In our previous work, we found that the observed pattern can be
explained from the hydrophobicity of the MHC-I binding cleft (\cite{bianchi2017}).
Here we do a similar analysis, by correlating the percentage of TMH-derived
epitopes presented versus the mean hydrophobicity of all presented
peptides. From these points, we derive a linear trendline and measure
the $R^2$, where $R^2$ is the proportion of the correlation explained by
the assumed linear relation.

@@

To determine if epitopes derived from TMHs are presented 
we used two elution studies, which both have sequenced
epitopes found in naturally occurring human immune cells,
for MHC-I (\cite{schellens2015comprehensive}) 
and MHC-II (\cite{bergseng2015different}).
For each of the sequenced epitopes,
we found its possible location(s) in 
a human representative reference proteome,
with UniProt ID UP000005640\_9606.
For the epitopes that were present only a single time in the proteome,
we predicted the topology of the proteins in which these epitopes are found.
We used both TMHMM and PureseqTM to do these predictions. 
From this topology, we could determine if the epitope
overlapped with a transmembrane region.

@@

The first query was a call to the 'gene' database for the 
term 'membrane protein' (in all fields) for the organism \emph{Homo sapiens}.
This resulted in 1,077 gene IDs.
% 1077 gene IDs is correct for December 2020.
% At 2021-03-01, one will get 1130 gene IDs.
% Also, one of the gene IDs that was valid back then,
% has been obsoleted.
The next query was a call to the 'gene' database 
to obtain the gene names from the gene IDs.
Per gene name, the 'dbSNP' NCBI database was queried for 
variations associated with the gene name. 
As the NCBI API constrains its users to three calls per second
to assure fair use, we had to limit the extent of our analysis.

The number of SNPs
was limited to the first 250 variations per gene (see figure
\ref{fig:snps_per_gene_name_ncbi} for the number of SNPs
within the dbSNP database for these gene names),
resulting in 60,683 variations at the protein level.
Only variations that result in a SNP for
a single amino acid substitution were analyzed, resulting in 38,882 SNPs.

@@

Figure \ref{fig:bbbq_1_smart_results} shows the predicted presentation of TMH-derived peptides in MHC-I,
for a human, viral and bacterial proteome.
Per MHC-I haplotype, it shows the percentage of binders that overlap with a TMH 
for at least one residue.
A dashed line indicates the expected percentage of TMH-derived epitopes 
that would be presented, if TMH-derived epitopes would be presented just as 
likely as soluble regions.
In 12 out of 13 MHC-I haplotypes, TMH-derived epitopes are presented more often 
then the null expectation, for a human proteome. 
For the viral and bacterial proteome, 11 out of 13 haplotypes present
TMH-derived epitopes more often than expected by chance.
The extent of the over-presentation between the various different haplotypes 
is similar for the probed proteomes, which strengthens our previous conclusion (\cite{bianchi2017}) that the hydrophobicity of the MHC-binding groove is the main factor responsible for the predicted over-presentation of TMH-derived peptides.

@@

  \begin{subfigure}[t]{0.4\textwidth}
    \centering
    \caption{}
    \includegraphics[width=\linewidth]{bbbq_1_smart_results/fig_hydrophobicity_mhc1.png}
    \label{fig:hydrophobicity_2}
  \end{subfigure}  
  \hfill
  \begin{subfigure}[t]{0.4\textwidth}
    \centering
    \caption{}
    \includegraphics[width=\linewidth]{bbbq_1_smart_results/fig_hydrophobicity_mhc2.png}
    \label{fig:hydrophobicity_2}
  \end{subfigure}  
  \caption{
    \textbf{(A)} 
    TMH-derived peptides are predicted to be over-presented in general,
    which is for MHC-I (left column) and MHC-II (right column),
    as well as for the proteomes of SARS-CoV-2 (top row), human (middle 
    row) and \emph{M. tuberculosis} (bottom row).
    The bars show the percentage of epitopes predicted to be 
    overlapping with TMHs.
    The red lines indicate the percentage as expected by chance.
    See supplementary tables \ref{tab:tmh_binders_mhc1} and \ref{tab:tmh_binders_mhc2}
    for the exact TMH and  epitope counts.
    \textbf{(B)} and \textbf{(C)}:
    The relation between the percentage of TMH-derived epitopes presented,
    and the hydrophobicity score of binders for MHC-I haplotypes (panel B)
    and MHC-II haplotypes (panel C). Trendline assumes a linear relationship.
  }
\end{figure}

@@

We next wondered if the over-representation of TMH-derived peptides would also be present for MHC-II. 
Figure \ref{fig:bbbq_1_smart_results} shows the percentages of MHC-II epitopes predicted to be overlapping 
with TMHs for our human, viral and bacterial proteomes.
See supplementary table \ref{tab:tmh_binders_mhc2} 
for the TMH and epitope counts.
We predict that TMH-derived peptides are also over-presented in most MHC-II haplotypes, 
and this over-presentation seems even more pronounced than for MHC-I.

This over-presentation can be explained by the hydrophobicity of the MHC-I binding cleft.
\richel{
  Elaborate.
  See \url{https://github.com/richelbilderbeek/bbbq_article/issues/192}
}
Similar to our previous results with MHC-I, we predicted that the extent to which TMH-derived
epitopes are presented by each MHC-II haplotype is similar for the human, bacterial and viral proteome.

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=0.5\textwidth]{bbbq_article_issue_157/results.png}
  \caption{
    TMH-derived epitopes are presented in vivo,
    for both MHC-I and MHC-II, for two topology prediction
    programs.
    See table \ref{tab:elution} for the exact values.
  }
  \label{fig:elution}
\end{figure}

To obtain experimental confirmation that peptides stemming from TMHs are presented in MHC-I and MHC-II,
peptide elution studies were reanalyzed.
Figure \ref{fig:elution} shows the percentages of epitopes derived 
from TMHs
found in the MHC-I (\cite{schellens2015comprehensive}) and MHC-II elution 
studies  (\cite{bergseng2015different}),
for the two topology prediction tools TMHMM (\cite{krogh2001predicting}) and PureseqTM (\cite{wang2019efficient}). 
Regardless of the prediction tool, 
at least 100 epitopes were predicted to be derived from a TMH for each condition. 
From these findings, it is robustly predicted that
epitopes derived from TMHs are presented in both MHC-I and MHC-II.

@@

  \caption{
    \textbf{(A)} 
    The number of SNPs in TMHs as expected by chance (left bar) 
    and found in the dbSNP database (right bar).
    Percentages are added to show the relative conservation
    of SNPs in TMHs found.
    \textbf{(B)}
    Percentage of SNPs found in TMHs.
    Each point shows the predicted percentage of
    TMH (x-axis) and the observed occurrence of SNPs being located
    within a TMH (y-axis) for a protein.
    The dashed diagonal line shows the line of equality (i.e.
    equal conservation of TMHs and soluble protein regions).
    The blue line is a linear trend line that includes only the
    transmembrane proteins.
    \textbf{(C)}
    Distribution of the percentages of TMH of TMPs.
  }
\end{figure}

@@

We addressed the question whether there is an evolutionary advantage in presenting TMHs. We determined the conservation of TMHs by comparing the occurrences of SNPs located in TMHs or solvent-exposed protein regions for the genes coding for membrane proteins.
We obtained $>$1,000 gene names associated with the phrase 'membrane protein',
which are genes for both membrane-associated proteins (MAPs, no TMH) and 
transmembrane proteins (TMPs, at least one TMH).
These genes are linked to 4,780 protein isoforms, 
of which 2,553 are predicted to be TMPs and 
2,237 proteins are predicted to be MAPs.
We obtained 37,630 SNPs that resulted in an
amino acids substitution, of which 21,024 were located 
in predicted TMPs.
See tables \ref{tab:ncbi_counts_1} and \ref{tab:ncbi_counts_2} 
for the detailed numbers and distributions of SNPs.

@@

We determined the probability to find the observed amount
of SNPs in TMHs by chance, i.e. when assuming SNPs occur 
just as likely in soluble domains as in TMHs.
We used a binomial Poisson distribution, 
where the number of trails ($n$) equals the number of SNPs, 
which is 21,208. 
The probability of success for the $i$th TMP ($p\_i$), 
is the percentage of residues within a TMH per TMP. 
These probabilities are shown as a histogram 
in figure \ref{fig:f_tmh_ncbi}). 
The expected number of SNPs expected to be found in 
TMHs by chance equals $\sum{p} \approx 4,141$.
As we observed 3,903 SNPs in TMHs, 
we calculated the probability of having that amount or less successes.
We used the type I error cut-off value of $\alpha = 2.5\%$.
The chance to find, within TMHs, this amount or less SNPs 
equals $6.8208 \cdot 10^{-11}$.
We determined the relevance of this finding, by
calculating how much less SNPs are found in TMHs,
when compared to soluble regions, which is the
ratio between the number of SNPs found in TMHs
versus the number of SNPs as expected by chance.
In effect, per 1000 SNPs found in soluble protein domains, 
one finds 918 SNPs in TMHs,
as depicted in figure \ref{fig:conservation}.

@@

% \paragraph{Over-presentation in vivo in MHC-I and MHC-II is robust
We show that 1.4\% of epitopes that are naturally presented on MHC-I 
can be mapped to stem from a TMH. We also find that MHC-II presents epitopes derived from TMHs in vivo, as we find that 3.9-4.0\% of sequenced epitopes
can be uniquely traced onto a human reference 
proteome (also shown in table \ref{tab:elution}).

@@

Although elution studies and in silico predictions both suggest that TMH-derived epitopes are presented in MHC-I and MHC-II, the molecular mechanisms of how integral membrane proteins are processed for MHC presentation are largely unknown (\cite{bianchi2017}). Most prominently, the fundamental principles of how TMHs are extracted from the  hydrophobic lipid environment into the aqueous vacuolar lumen, and their prior or subsequent proteolytic processing are unresolved.
\geert{
  beetje kort. Suggesties? Experimental approaches? Hoeft niet super
  origineel, bijvoorbeeld suggereren dat lipases een rol kunnen spelen. Frans kan vast helpen met mooie zin
}

@@

An important caveat of the elution studies is that we assume that 
the found epitopes are derived from their completely 
translated predecessor proteins.
However, there are multiple cellular processes 
that may result in peptide fragments that did not originate 
from the mature proteins.
One such process is the production of defective ribosomal 
products (DRiPs) (\cite{yewdell1996defective}),
in which the ribosome terminates its translation prematurely,
intrudes a point mutation, 
or cause a frame shift, at low frequencies (\cite{yewdell2011drips}).
Another process that results in peptide fragments absent in the
proteome is peptide fusion, 
in which peptides form covalent bonds with other peptides,
as is for example found in mice (\cite{delong2016pathogenic}).
Due to these processes, we may incorrectly attribute an
epitope to a specific proteome locations, 
where in reality, it was created de novo.
This may lead to false positives, as we incorrectly deduce
epitopes to be derived from a TMH, instead of from one of
these processes. Both processes, however, are thought to
occur at low frequencies (\cite{yewdell2011drips,delong2016pathogenic}).

@@

Not only do we predict that MHC-I and MHC-II TMH-derived epitopes 
are presented, we conclude that these are additionally
presented more often than expected by chance 
alone (figure \ref{fig:bbbq_1_smart_results}).
This in silico finding is a general pattern 
that persists in both MHC-I and MHC-II, 
for humans, bacteria and viruses.

% \paragraph{MHC-II is expected to present TMHs}
The results from this study support our hypothesis 
that TMH-derived epitopes are over-presented
in MHC-II, similarly to we previously showed for MHC-I (\cite{bianchi2017}). 
The over-presentation of TMHs for most MHC-II haplotypes 
appears to follow a similar pattern as MHC-I haplotypes,
although MHC-II presents slightly more TMH-derived
epitopes (see Supplementary Materials, figure \ref{fig:rel_presentation}
for a direct comparison). 
Unlike MHC-I, the human proteome over-presents TMH-derived
epitopes for all MHC-II haplotypes that we have investigated.
Possible there are MHC-II haplotypes that present less TMH-derived
epitopes than expected by chance.
The haplotype \verb;HLA-DQA1*0101/DQB1*0501; had the highest
over-presentation of TMH-derived epitopes,
which we suggest is the haplotype of choice 
to investigate the intracellular mechanism of TMH-derived
antigen presentation.

>>>>

Cytolytic T cell responses are predicted to be biased towards membrane proteins. 
Because the peptide-binding grooves of most haplotypes 
of histocompatibility complex class I (MHC-I) are relatively hydrophobic, 
peptide fragments derived from human transmembrane helices (TMHs) 
are predicted to be presented more often than expected,
based on their abundance in the proteome.
We show that this over-presentation of TMH-derived peptides is general, 
as it is predicted for diverse bacteria and viruses 
and for both MHC-I and MHC-II.  
However, the physiological reason of why membrane proteins might be 
over-presented is unclear
In this study, we show that TMHs are evolutionarily more conserved, 
as single nucleotide polymorphisms (SNPs) 
are present relatively less frequent in TMH-coding chromosomal regions 
compared to regions coding for extracellular and cytosolic protein regions. 
Thus, our findings suggest that T cells might respond 
more to membrane proteins, because these are evolutionary more conserved.
We speculate that TMHs therefor might be less prone to escape mutations 
that enable pathogens to evade T cell responses.

@@

The MHC proteins are heterodimeric proteins encoded by the
HLA (Human Leukocyte Antigens) genes.
In humans, the peptide-binding groove of MHC-I is made by the alpha chain, 
and there are three forms of alpha chains called HLA-A, HLA-B and HLA-C. 
For MHC-II, both the alpha and the beta chains contribute to the binding groove, 
and there are three major forms as well, called HLA-DR, HLA-DQ and HLA-DP.
Each MHC complex can only bind a subset of all possible peptides.
For example, HLA-A and HLA-B have no overlap in which
peptides they bind (\cite{lund2004definition}).
The HLA genes of humans are highly polymorphic, with hundreds 
to thousands of different alleles, 
and each different HLA allele is called 
an MHC haplotype (\cite{marsh2010nomenclature}).

@@

While humans only express a limited number of MHC haplotypes, 
and therefore any individual's immune system detects 
only a fraction of all possible peptide fragments, 
the number of pathogenic peptides that can be detected 
is increased at the population level, 
because of the multiple and highly polymorphic MHC genes.
It is believed that this improves immunity at the population level, 
as mutations in a protein that disrupt a particular MHC presentation, 
so-called escape mutations, 
will not affect MHC presentation for all haplotypes (\cite{sommer2005importance}).

@@

Many studies are aimed at determining which peptides are presented in MHC 
and will result in an immune response, 
as this will for instance aid the design of vaccines. 
These studies have led to the development of prediction algorithms 
that allow for very reliable \emph{in silico} predictions 
of the binding affinities of peptides
(\cite{larsen2010identification,schellens2008unanticipated,tang2011genome}).
For example, \cite{tang2011genome} found that, 
of the 432 peptides that were predicted to bind to MHC,
86\% were confirmed to do so in vivo.

@@

TMHs are hydrophobic 
as they need to span the hydrophobic lipid bilayer of cellular membranes.
They consist of an alpha helix of, on average, 23 amino acids in length. 
TMHs can also be predicted from a protein sequence 
with high accuracy by bioinformatics approaches (\cite{krogh2001predicting,bianchi2017,kall2004combined,arai2004conpred,jones2007improving,klammer2009metatm,wang2019efficient}),
for example, \cite{jones2007improving} found that,
from 184 TMPs with known topology, 
80\% of the TMH predictions replicated this finding.

TMHs are common structures in the proteins of humans and microbes. 
Different TMH prediction tools estimate that 15-39\% of all proteins 
in the human proteome contain at least one TMH (\cite{ahram2006estimation}).
However, the physiological reason why peptides derived from TMHs 
would be presented more often than peptides 
stemming from soluble protein regions is unknown. 
We hypothesized that the presentation of 
TMH residues is evolutionary selected for, 
because TMHs are less prone to undergo escape mutation. 
One reason to expect such a reduced 
variability (and hence evolutionary conservation) in TMHs, 
is that these are restricted in their evolution 
by the functional requirement to span a lipid bilayer. 
Due to this requirement, 
many of the amino acids genuinely present in TMHs 
are limited to the ones with hydrophobic side chains (\cite{jones1994model}).
Therefore, we speculated that the TMHs of pathogens 
might have a lower chance to develop escape mutations, 
as many mutations will result in a dysfunctional TMH 
and render the protein inactive.

This study has two objectives. 
First, we aimed to generalize our findings by predicting the presentation of peptides 
from different kingdoms of life and for both MHC-I and -II. 
From these \emph{in silico} predictions, we conclude that TMH-derived
epitopes are presented more often than expected by chance,
in a human, viral and bacterial proteome, 
and for most haplotypes of both MHC-I and II. 
We confirmed this over-presentation of TMH-derived peptides 
by re-analysis of a peptide elution study. 
Second, we tested our hypothesis that TMHs 
are more evolutionarily conserved than solvent-exposed protein regions. 
Our analysis of human single nucleotide polymorphisms (SNPs) showed 
that random point mutations are indeed less likely
to occur within TMHs. 
These findings strengthen the emerging notion 
that TMHs are important for the T cell branch of the adaptive immune system, 
and hence are of  
overlooked importance in vaccine development.

@@

The MHC-I molecule presents epitopes of shorter length than MHC-II.
We expect this to be of no influence in the presentation level
of TMH-derived epitopes, for each of the three proteomes.
We verified that MHC-II is just as prone to over-present TMH-derived
epitopes by comparing the relative over-presentation of these epitopes.

In our previous work, we found that the observed pattern can be
explained from the hydrophobicity of the MHC-I binding cleft (\cite{bianchi2017}).
Here we do a similar analysis, by correlating the percentage of TMH-derived
epitopes presented versus the mean hydrophobicity of all presented
peptides. From these points, we derive a linear trendline and measure
the proportion of the correlation explained by the assumed linear relation.

@@

To determine if epitopes derived from TMHs are presented \emph{in vivo}
we used two elution studies, both of which have sequenced 
the epitopes found in naturally occurring human immune cells,
for MHC-I (\cite{schellens2015comprehensive}) 
and MHC-II (\cite{bergseng2015different}).
For each of the sequenced epitopes,
we found its possible location(s) in 
a human representative reference proteome,
with UniProt ID UP000005640\_9606.
For the epitopes that were present in the proteome exectly once,
we predicted the topology of the proteins in which these epitopes are found.
We used both TMHMM (\cite{krogh2001predicting}) and PureseqTM (\cite{wang2019efficient}) to do these predictions. 
From this topology, we could determine if the epitope
overlapped with a transmembrane region.

@@

The first query was a call to the 'gene' database for the 
term 'membrane protein' (in all fields) 
for the organism \emph{Homo sapiens}.
This resulted in 1,077 gene IDs (on December 2020).
% 1077 gene IDs is correct for December 2020.
% At 2021-03-01, one will get 1130 gene IDs.
% Also, one of the gene IDs that was valid back then,
% has been obsoleted.
The next query was a call to the 'gene' database 
to obtain the gene names from the gene IDs.
Per gene name, the 'dbSNP' NCBI database was queried for 
variations associated with the gene name. 
As the NCBI API constrains its users to three calls per second
(to assure fair use), we had to limit the extent of our analysis.

The number of SNPs was limited to the first 250 variations per gene,
resulting in $\approx$61k variations.
Only variations that result in a SNP for
a single amino acid substitution were analyzed, resulting in $\approx$38k SNPs.
The exact amounts can be found in the supplementary materials,
tables \ref{tab:ncbi_counts_1} and \ref{tab:ncbi_counts_2}.

@@

Figure \ref{fig:bbbq_1_smart_results} shows the predicted presentation of TMH-derived peptides in MHC-I,
for a human, viral and bacterial proteome.
Per MHC-I haplotype, it shows the percentage of binders that overlap with a TMH 
for at least one residue.
A horizontal line shows the expected percentage of TMH-derived epitopes 
that would be presented, if TMH-derived epitopes would be presented just as 
likely as soluble regions.
In 11 out of 13 MHC-I haplotypes, TMH-derived epitopes are presented more often 
then the null expectation, for a human and bacterial proteome.
For the viral proteome, 12 out of 13 haplotypes present
TMH-derived epitopes more often than expected by chance.
The extent of the over-presentation between the various different haplotypes 
is similar for the probed proteomes, 
which strengthens our previous conclusion (\cite{bianchi2017}) 
that the hydrophobicity of the MHC-binding groove 
is the main factor responsible for the predicted over-presentation 
of TMH-derived peptides.

@@

  \begin{subfigure}[t]{0.4\textwidth}
    \centering
    \caption{}
    \includegraphics[width=\linewidth]{bbbq_1_smart_results/fig_hydrophobicity_mhc1.png}
    \label{fig:hydrophobicity_1}
  \end{subfigure}  
  \hfill
  \begin{subfigure}[t]{0.4\textwidth}
    \centering
    \caption{}
    \includegraphics[width=\linewidth]{bbbq_1_smart_results/fig_hydrophobicity_mhc2.png}
    \label{fig:hydrophobicity_2}
  \end{subfigure}  
  \caption{
    \textbf{(A)} 
    TMH-derived peptides are predicted to be over-presented in general,
    which is for MHC-I (left column) and MHC-II (right column),
    as well as for the proteomes of SARS-CoV-2 (top row), human (middle 
    row) and \emph{M. tuberculosis} (bottom row).
    The bars show the percentage of epitopes predicted to be 
    overlapping with TMHs.
    The red lines indicate the percentage as expected by chance.
    See supplementary tables \ref{tab:tmh_binders_mhc1} and \ref{tab:tmh_binders_mhc2}
    for the exact TMH and  epitope counts.
    \textbf{(B)} and \textbf{(C)}:
    The relation between the percentage of TMH-derived epitopes presented,
    and the hydrophobicity score of binders for MHC-I haplotypes (panel B)
    and MHC-II haplotypes (panel C). The trendline assumes a linear relationship.
  }
\end{figure}

@@

We next wondered if the over-representation of TMH-derived peptides 
would also be present for MHC-II. 
Figure \ref{fig:bbbq_1_smart_results} shows the percentages of MHC-II epitopes 
predicted to be overlapping with TMHs for our human, viral and bacterial proteomes.
We found that TMH-derived peptides are over-presented in all
of the 21 MHC-II haplotypes, 
in a human, bacterial and viral proteome,
except for \verb;HLA-DRB3*0101; in \emph{MTb}
See supplementary table \ref{tab:tmh_binders_mhc2} 
for the exact TMH and epitope counts.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Hydrophocity explains the generality of this finding}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Hydrophobicity may be the common reason of the consistent over-presentation
of  (hydrophobic) TMH-derived epitopes.
Figures \label{fig:hydrophobicity_1} and \label{fig:hydrophobicity_2}
show the hydrophobicity preference score for the extent over-presentation
of TMH-derived epitopes for the different haplotypes in a human proteome.
An assumed linear correlation explains 88\% of the variability in MHC-I.
For MHC-II, 62\% of the variability is explained by hydrophobicity.

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=0.5\textwidth]{bbbq_article_issue_157/results.png}
  \caption{
    TMH-derived epitopes are presented \emph{in vivo},
    for both MHC-I and MHC-II, for two topology prediction
    programs.
  }
  \label{fig:elution}
\end{figure}

To obtain experimental confirmation that peptides stemming from TMHs 
are presented in MHC-I and MHC-II,
two peptide elution studies (\cite{schellens2015comprehensive} and \cite{bergseng2015different} for MHC-I and MHC-II respectively) 
were reanalyzed.
Figure \ref{fig:elution} shows the percentages of epitopes derived 
from TMHs
found in the MHC-I and MHC-II elution 
studies,
for the two topology prediction tools TMHMM (\cite{krogh2001predicting}) and PureseqTM (\cite{wang2019efficient}). 
Regardless of the prediction tool, 
at least 100 epitopes were predicted to be derived from a TMH for each condition. 
From these findings, it is robustly predicted that
epitopes derived from TMHs are presented in both MHC-I and MHC-II.
See the supplementary materials (table \ref{tab:elution}) for the exact values.

@@

  \caption{
    \textbf{(A)} 
    The number of SNPs in TMHs as expected by chance (left bar) 
    and found in the dbSNP database (right bar).
    Percentages are added to show the relative conservation
    of SNPs in TMHs found.
    \textbf{(B)}
    Percentage of SNPs found in TMHs.
    Each point shows the predicted percentage of
    TMH (x-axis) and the observed occurrence of SNPs being located
    within a TMH (y-axis) for a protein.
    The dashed diagonal line shows the line of equality (i.e.
    equal conservation of TMHs and soluble protein regions).
    \textbf{(C)}
    Distribution of the percentages of TMH in the TMPs used in this study.
  }
\end{figure}

@@

We addressed the question whether there is an evolutionary advantage in presenting TMHs. 
We determined the conservation of TMHs 
by comparing the occurrences of SNPs located in TMHs or solvent-exposed protein regions 
for the genes coding for membrane proteins.
We obtained 911 unique gene names associated with the phrase 'membrane protein',
which are genes for both membrane-associated proteins (MAPs, which have no TMH) and 
transmembrane proteins (TMPs, which have at least one TMH).
These genes are linked to 4,780 protein isoforms, 
of which 2,553 are predicted to be TMPs and 
2,237 proteins are predicted to be MAPs.
We obtained 37,630 unique variations, 
of which 9,621 are SNPs that resulted in a straightforward amino acids substitution, 
of which 6,062 were located in predicted TMPs.
See tables \ref{tab:ncbi_counts_1} and \ref{tab:ncbi_counts_2} 
for the detailed numbers and distributions of SNPs.

@@

We determined the probability to find the observed amount
of SNPs in TMHs by chance, i.e. when assuming SNPs occur 
just as likely in soluble domains as in TMHs.
We used a binomial Poisson distribution, 
where the number of trails ($n$) equals the number of SNPs, 
which is 21,208. 
The probability of success for the $i$th TMP ($p\_i$), 
is the percentage of residues within a TMH per TMP. 
These probabilities are shown as a histogram 
in figure \ref{fig:f_tmh_ncbi}). 
The expected number of SNPs expected to be found in 
TMHs by chance equals $\sum{p} \approx 4,141$.
As we observed 3,803 SNPs in TMHs, 
we calculated the probability of having that amount or less successes.
We used the type I error cut-off value of $\alpha = 2.5\%$.
The chance to find, within TMHs, this amount or less SNPs 
equals $6.8208 \cdot 10^{-11}$.
We determined the relevance of this finding, by
calculating how much less SNPs are found in TMHs,
when compared to soluble regions, which is the
ratio between the number of SNPs found in TMHs
versus the number of SNPs as expected by chance.
In effect, per 1000 SNPs found in soluble protein domains, 
one finds 918 SNPs in TMHs,
as depicted in figure \ref{fig:conservation}.

@@

% \paragraph{Over-presentation in vivo in MHC-I and MHC-II is robust
We show that 1.4\% of epitopes that are naturally presented on MHC-I 
can be mapped to stem from a TMH. 
We also find that MHC-II presents epitopes derived from TMHs \emph{in vivo}, 
as we find that 3.9-4.0\% of sequenced epitopes
can be uniquely traced onto a human reference proteome (also shown in table \ref{tab:elution}).

@@

Although elution studies and \emph{in silico} predictions both suggest that 
TMH-derived epitopes are presented in MHC-I and MHC-II, 
the molecular mechanisms of how integral membrane proteins are processed 
for MHC presentation are largely unknown (\cite{bianchi2017}). 
Most prominently, the fundamental principles of 
how TMHs are extracted from the hydrophobic lipid environment 
into the aqueous vacuolar lumen, 
and their prior or subsequent proteolytic processing are unresolved.
\geert{
  beetje kort. Suggesties? Experimental approaches? Hoeft niet super
  origineel, bijvoorbeeld suggereren dat lipases een rol kunnen spelen. Frans kan vast helpen met mooie zin
}

@@

An important caveat of the elution studies is that we assume that 
the found epitopes are derived from their completely 
translated predecessor proteins.
However, there are multiple cellular processes 
that may result in peptide fragments that did not originate 
from the mature proteins.
One such process is the production of defective ribosomal 
products (DRiPs) (\cite{yewdell1996defective}),
in which the ribosome terminates its translation prematurely,
or cause a point mutation or frame shift (\cite{yewdell2011drips}).
Another process that results in peptide fragments absent in the
proteome is peptide fusion, 
in which peptides form covalent bonds with other peptides,
as is for example found in mice (\cite{delong2016pathogenic}).
Due to these processes, we may incorrectly attribute an
epitope to a specific proteome locations, 
where in reality, it was created \emph{de novo}.
This may lead to false positives, as we incorrectly deduce
epitopes to be derived from a TMH, instead of from one of
these processes. Both processes, however, are thought to
occur at low frequencies (\cite{yewdell2011drips,delong2016pathogenic}).

@@

Not only do we predict that MHC-I and MHC-II TMH-derived epitopes 
are presented, we conclude that these are additionally
presented more often than expected by chance 
alone (figure \ref{fig:bbbq_1_smart_results}).
This \emph{in silico} finding is a general pattern 
that persists in both MHC-I and MHC-II, 
for humans, bacteria and viruses.

% \paragraph{MHC-II is expected to present TMHs}
The results from this study support our hypothesis 
that TMH-derived epitopes are over-presented
in MHC-II, similar to what we previously showed for MHC-I (\cite{bianchi2017}). 
The over-presentation of TMHs for most MHC-II haplotypes 
appears to follow a similar pattern as MHC-I haplotypes,
although MHC-II presents slightly more TMH-derived
epitopes (see Supplementary Materials, figure \ref{fig:rel_presentation}
for a direct comparison). 
Unlike MHC-I, the human proteome over-presents TMH-derived
epitopes for all MHC-II haplotypes that we have investigated.
Possible there are MHC-II haplotypes that present less TMH-derived
epitopes than expected by chance.
The haplotype \verb;HLA-DQA1*0101/DQB1*0501; had the highest
over-presentation of TMH-derived epitopes,
which we suggest is the haplotype of choice 
to investigate the intracellular mechanism of TMH-derived
antigen presentation.
